Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-10 paper.tex
Para on Zarr's limitations

Table~\ref{tab-genomics-england-data} shows that the dataset storage
size is dominated by a few columns with the top four
(call\_AD, call\_GQ, call\_DP and call\_DPF) accounting for over 
80\% of the total. These fields are much less compressible
than genotype data (which uses $<1\%$ of the total space here)
because of their inherent noisiness~\citep{lin2020sparse}. By 
default the \texttt{bio2zarr} conversion is lossless, and integers
of the narrowest type required to represent all of the data are
used. In this case, this has resulted in 16 bit integers being used 
for these top four fields. Examining the distribution of the depth
fields, (call\_AD, call\_DP, call\_DPF), we can 
we can see that a tiny fraction of the values stored
are greater than 127 (call\_AD: 0.1\%, call\_DP: xx\%, call\_DPF: x\%; see 
Fig SX for histograms).
One might then reasonably truncate these fields to 8 bit integers.
After converting to 8 bit, call\_AD is reduced by 
x\% (XX GiB) and call\_DP reduced by x\% (XX GiB)
and call\_DPF by x\% (XX GiB). 
Similarly, the vast majority of call\_GQ values (xx\%) are less than 
128 (corresponding to a conditional probability that the genotype call is 
wrong of $10^{-128}$ [TODO CHECK THIS]), and can also reasonably
be truncated. Here, this results in a reduction of x\% (XX GiB).
Combined, truncating these top four fields to 8 bits reduces the 
overall storage of the dataset to XXGiB, corresponding to a 
X-fold compression of the VCF.

@@

Calculating the histogram of per call values at this 
scale is not a trivial task.
Taking advantage of the rich software ecosystem growing around Zarr
(see Discussion) we used xarray~\citep{hoyer2017xarray}
and Dask~\citep{rocklin2015dask} to transparently distribute the 
computations over a cluster of X workers. 
[ADD DETAILS]
The elapsed time was X minutes, and requires X lines of Python in a 
Jupyter notebook~\citep{kluyver2016jupyter}.

One of the key uses of the aggV2 dataset is to enable simple queries 
of specific regions. For example, in [A PAPER] they [DID SOMETHING].
We replicated this by running the appropriate bcftools query
on the VCF chunk (note that the user is responsible for finding 
the correct chunk file here, limiting the utility of genomic 
indexes). This took XXX hours.
The same query implemented with the Zarr-python 
API required XXX seconds to produce identical output.

\section{Discussion}
% The introduction of VCF as part of the work of the 1000 Genomes
% project~\cite{10002015global} has transformed a fragmented 
% landscape of mutually incompatible per-program file formats
% into an ecosystem of tools that largely
% interoperate~\cite{garrison2022spectrum}. In this 
% sense VCF has been a huge success, and the underlying data model
% is now deeply embedded in bioinformatics practice. As we have 
% illustrated here, the shortcomings of the VCF and BCF file
% formats (and more generally \emph{any} row-wise encoding of the 
% data model) in terms of computational accessibility are severe,
% and there can be little doubt that these file formats 
% are incurring a large economic cost and 
% holding back progress. 
% As we have illustrated in the Genomics England
% example here, these shortcomings are most acutely felt in 
% large-scale data human datasets but any setting in which we 
% wish to extract columns of the data from a row-wise 
% VCF format will be inefficient 
% when datasets are larger than a few hundred megabytes.

% As we have demonstrated, the columnar approach, based on the 
% popular Zarr standard and widely used compression libraries, 
% can support many different types of data access
% patterm efficiently, and has excellent computational accessibility.
% In this approach there are some significant departures from 
% standard practice, which warrant some discussion here.

@@

% Probably getting into the weeds too much.
% Some large cloud-hosted VCF datasets are currently stored in 
% filesystems that emulate POSIX semantics on top of such object 
% stores, at a substantial cost in performance. Directly hosting 
% these datasets in object stores would have substantial 
% benefits, both in terms of storage costs and performance.
% Thirdly, breaking up fields into different directories (or 
% cloud buckets) enables fine-grained access control to the 
% dataset. For example, all users may be allowed access to
% variant-level data, but only a subset permitted to access
% call-level individual data. Such access controls can be 
% straightforwardly applied using file or cloud-hosting 
% methods on the relevant directores/buckets. Such concerns
% are entirely decoupled from the file format.

% This is poorly explained
Along with the move away from the single-file-storage paradigm,
the approach we outline here also suggests a shift from 
the classical Unix pipeline-based processing approach 
(but see below for a discussion of how existing pipelines can
be supported). Datasets flowing through pipelines inherently 
implies creating muliple copies of the data, and when 
multiple terabytes of either unrelated or unchanged data 
get copied several times through these pipes large inefficiencies 
are incurred (see Fig~\ref{fig-whole-matrix-compute}). 
We may think of this strategy as ``copy-oriented'' workflows.
A more efficient way to structure such multilayered computations
is in a  ``mask-oriented'' way. For example, to implement some
complex multilayered call filtering strategy we could imagine sequentially
updating a given per-call mask. Masks are cheap (e.g., the 
standard genotype missingness mask in Table~\ref{tab-genomics-england-data}
uses less than 1\% of the overall storage).

Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a solution to all problems. [large number
of alleles, dataset is basically static, ragged arrays,
strings ... any others? Zarr v3 on the way.]

@@

% Zarr has also seen recent uptake 
% in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
% and multimodal spatial omics
% data~\citep{marconato2024spatialdata,baker2023emobject}.
% Recent additions using Zarr include the application of deep learning
% models to genomic sequence data~\citep{klie2023predictive} 
% and a browser for genetic variation data~\citep{konig2023divbrowse}.

>>>>

Table~\ref{tab-genomics-england-data} shows that the dataset storage
size is dominated by a few columns with the top four
(call\_AD, call\_GQ, call\_DP and call\_DPF) accounting for over 
80\% of the total. These fields are much less compressible
than genotype data (which uses $<1\%$ of the total space here)
because of their inherent noisiness~\citep{lin2020sparse}.

[TODO add in benchmarks discussed in issue. Extract POS, run af-dist, 
do a filtering query. Note problems with orchestration when
dealing with multiple VCF chunks. 
Talk about mask-oriented computing there at the
end, pointing out the cost of these multiple copies.]

% By default the \texttt{bio2zarr} conversion is lossless, and integers
% of the narrowest type required to represent all of the data are
% used. In this case, this has resulted in 16 bit integers being used 
% for these top four fields. Examining the distribution of the depth
% fields, (call\_AD, call\_DP, call\_DPF), we can 
% we can see that a tiny fraction of the values stored
% are greater than 127 (call\_AD: 0.1\%, call\_DP: xx\%, call\_DPF: x\%; see 
% Fig SX for histograms).
% One might then reasonably truncate these fields to 8 bit integers.
% After converting to 8 bit, call\_AD is reduced by 
% x\% (XX GiB) and call\_DP reduced by x\% (XX GiB)
% and call\_DPF by x\% (XX GiB). 
% Similarly, the vast majority of call\_GQ values (xx\%) are less than 
% 128 (corresponding to a conditional probability that the genotype call is 
% wrong of $10^{-128}$ [TODO CHECK THIS]), and can also reasonably
% be truncated. Here, this results in a reduction of x\% (XX GiB).
% Combined, truncating these top four fields to 8 bits reduces the 
% overall storage of the dataset to XXGiB, corresponding to a 
% X-fold compression of the VCF.

@@

% Calculating the histogram of per call values at this 
% scale is not a trivial task.
% Taking advantage of the rich software ecosystem growing around Zarr
% (see Discussion) we used xarray~\citep{hoyer2017xarray}
% and Dask~\citep{rocklin2015dask} to transparently distribute the 
% computations over a cluster of X workers. 
% [ADD DETAILS]
% The elapsed time was X minutes, and requires X lines of Python in a 
% Jupyter notebook~\citep{kluyver2016jupyter}.

% One of the key uses of the aggV2 dataset is to enable simple queries 
% of specific regions. For example, in [A PAPER] they [DID SOMETHING].
% We replicated this by running the appropriate bcftools query
% on the VCF chunk (note that the user is responsible for finding 
% the correct chunk file here, limiting the utility of genomic 
% indexes). This took XXX hours.
% The same query implemented with the Zarr-python 
% API required XXX seconds to produce identical output.

\section{Discussion}
VCF is a central element of modern genomics, facilitating
the exchange of data among interoperating tools. 
Its current row-oriented form, however,
is fundamentally inefficient,
profoundly limiting the scalability of the current generation
of bioinformatics tools. Large scale VCF data cannot 
currently be 
processed without incurring a substantial economic
and likely environmental cost~\cite{grealey2022carbon}.
We have shown here that this is not a necessary situation,
and that greatly improved efficiency can be achieved by 
using more appropriate storage representations tuned
to the realities of modern computing. We have argued that 
Zarr provides a powerful basis for cloud-based
storage and analysis of large-scale genetic variation data.
We propose the VCF Zarr specification which losslessly
maps VCF data to Zarr, and provide an efficient and scalable
tool to perform conversion.

Zarr provides pragmatic solutions to some of the more pressing 
problems facing the analysis of large-scale genetic variation
data, but it is not a solution to all problems. Firstly, 
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the 
dimensions of various fields are determined by the \emph{maximum}
dimension among all variants. Secondly, the design of 
VCF Zarr emphasises efficiency of analysis for a fixed 
dataset, and does not consider problems of ongoing updates
(i.e, as new samples and their variants are added).
Thirdly, Zarr is not suitable for handling unstructured
string data, and for example likely not work well as a 
basis for storing annotation data [CITES].

@@

@@
